\documentclass[12pt]{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}

\usepackage{setspace}
\onehalfspacing

\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm}
\usepackage{epsfig}
\usepackage{times}
\renewcommand{\ttdefault}{cmtt}
\usepackage{amsmath}
\usepackage{graphicx} % for graphics files

% Draw figures yourself
\usepackage{tikz} 

% writing elements
\usepackage{mhchem}

% The float package HAS to load before hyperref
\usepackage{float} % for psuedocode formatting
\usepackage{xspace}

% from Denovo Methods Manual
\usepackage{mathrsfs}
\usepackage[mathcal]{euscript}
\usepackage{color}
\usepackage{array}

\usepackage[pdftex]{hyperref}
\usepackage[parfill]{parskip}

% math syntax
\newcommand{\nth}{n\ensuremath{^{\text{th}}} }
\newcommand{\ve}[1]{\ensuremath{\mathbf{#1}}}
\newcommand{\Macro}{\ensuremath{\Sigma}}
\newcommand{\rvec}{\ensuremath{\vec{r}}}
\newcommand{\vecr}{\ensuremath{\vec{r}}}
\newcommand{\omvec}{\ensuremath{\hat{\Omega}}}
\newcommand{\vOmega}{\ensuremath{\hat{\Omega}}}
\newcommand{\sigs}{\ensuremath{\Sigma_s(\rvec,E'\rightarrow E,\omvec'\rightarrow\omvec)}}
\newcommand{\el}{\ensuremath{\ell}}
\newcommand{\sigso}{\ensuremath{\Sigma_{s,0}}}
\newcommand{\sigsi}{\ensuremath{\Sigma_{s,1}}}
%---------------------------------------------------------------------------
%---------------------------------------------------------------------------
\begin{document}
\begin{center}
{\bf NE 250, F15\\
October 14, 2015 
}
\end{center}

The \textbf{Adjoint Transport Equation} (L\&M 1.6).

Last time we started talking about the adjoint transport equation and began be defining inner products. 

The \textit{adjoint operator} obeys the following identity: $\langle A^{\dagger}f, g\rangle = \langle f, Ag\rangle$ for all $f$ and $g$ is the vector space. Superscript $\dagger$ indicates adjoint quantities.\\
Note that if $M_1^{\dagger}$ and $M_2^{\dagger}$ are both adjoints of $M_1$ and $M_2$, respectively, then $(M_1 + M_2)^{\dagger}  = M_1^{\dagger} + M_2^{\dagger}$.

We can relate the forward and adjoint eigenspaces\\
Forward: $Mf = \lambda f$ with homogeneous BCs;\\
Adjoint: $M^{\dagger}f^{\dagger} = \lambda^{\dagger} f^{\dagger}$ with homogeneous BCs.

\begin{align*}
\langle M^{\dagger}f^{\dagger}, f\rangle - \langle f^{\dagger}, Mf\rangle &= \langle\lambda^{\dagger} f^{\dagger}, f\rangle - \langle f^{\dagger} ,\lambda f\rangle \\
0 &= \langle\lambda^{\dagger} - \lambda\rangle \langle f^{\dagger}, f\rangle \quad \text{orthogonality condition}
\end{align*}
%
If $\lambda^{\dagger, *} \neq \lambda$ then $\langle f^{\dagger}, f\rangle = 0$.\\
If $\langle f^{\dagger}, f\rangle \neq 0$ then $\lambda^{\dagger, *} = \lambda \rightarrow \lambda^{\dagger} = \lambda$.\\
If $M = M^{\dagger}$, the operator is called self adjoint. 

The \textit{Adjoint Transport Equation for non-multiplying media}.

Note that we're going to use $\zeta$ to indicate this could be applied to any appropriate function (though in practice we'll use $\psi$).

We will use vacuum boundary conditions: $\zeta(\vec{r}_s, \vOmega, E) = 0$, $\hat{n} \cdot \vOmega < 0$, where $\vec{r}_s$ is on the surface and $\hat{n}$ is the outward normal at $\vec{r}_s$.\\
Recall that $\zeta(\vec{r}_s, \vOmega, E)$ must be continuous in the direction of particle travel.


We will can $H$ our transport operator.\\
Our goal is to get something that looks like $\langle\zeta^{\dagger}, H \zeta\rangle = \langle\zeta, H^{\dagger} \zeta^{\dagger}\rangle$.\\


To do get the lhs, we will multiply the regular TE by $\zeta^{\dagger}$, ingrate over all phase space, and manipulate the result to get $\zeta$ in front of the operator.
%
\[\int_{\rho} d\rho\: \zeta^{\dagger}(\rho)\biggl[ \bigl[\vOmega \cdot \nabla + \Sigma_t\bigr] \zeta(\vec{r}, E, \vOmega) - \int_{4 \pi} d\vOmega' \int_0^{\infty} dE' \: \Sigma_s(E', \vOmega' \rightarrow E, \vOmega) \zeta(\vec{r}, E', \vOmega') \biggr]\]
% 
To look at one term at a time:
\begin{itemize}
\item The total interaction is a self-adjoint operator, so we can simply rearrange the order: $\zeta^{\dagger} \Sigma_t \zeta \rightarrow \zeta \Sigma_t \zeta^{\dagger}$. That is
\[\int_V dV \zeta^{\dagger}(\vec{r}, E, \vOmega) \Sigma_t(E) \zeta(\vec{r}, E, \vOmega) = \int_V dV \zeta(\vec{r}, E, \vOmega) \Sigma_t(\rvec,E) \zeta^{\dagger}(\vec{r}, E, \vOmega)\]
%
\item For the streaming term, use $\vOmega \cdot \nabla = \nabla \cdot \vOmega$ and the identity (from the chain rule)
\[\nabla \cdot \bigl(\vOmega \zeta^{\dagger} \zeta \bigr) = \zeta^{\dagger}\vOmega \cdot \nabla \zeta + \zeta \vOmega \cdot \nabla \zeta^{\dagger}\]
then
\[
\int_V d^3r \:\zeta^{\dagger}\vOmega \cdot \nabla \zeta + \int_V d^3r \:\zeta \vOmega \cdot \nabla \zeta^{\dagger} = \int_S d \Gamma \hat{n} \cdot \vOmega \zeta^{\dagger} \zeta
\]
from divergence theorem. Now, we apply the boundary condition that we originally had for $\zeta$, which causes the surface integral to vanish for $\vOmega \cdot \hat{n} < 0$.

We have a corresponding boundary condition for $\zeta^{\dagger}$ that at any point $\rvec_s$ on the surface, $\zeta^{\dagger}(\rvec, \vOmega, E) = 0$, $\vec{r} \in \Gamma$ and $\vOmega \cdot \hat{n} \geq 0$.

This means that the surface integral has to vanish for all $\vOmega$ giving
\[
\int_V d^3r \:\zeta^{\dagger}\vOmega \cdot \nabla \zeta = -\int_V d^3r \:\zeta \vOmega \cdot \nabla \zeta^{\dagger}
\]
%
\item Finally we'll look at the scattering term. As with total interaction, we know $\zeta^{\dagger} \Sigma_s \zeta  = \zeta \Sigma_s \zeta^{\dagger}$. \\
We will also swap $E'$ and $E$, $\vOmega'$ and $\vOmega$ to get (leaving out volume integration but we'll add it later)
\begin{align*}
\int d\vOmega \int dE\: &\zeta^{\dagger}(\rvec, \vOmega, E) \int d\vOmega' \int dE'\: \Sigma_s(\rvec, E' \rightarrow E, \vOmega' \cdot \vOmega) \zeta(\rvec, \vOmega', E') = \\
%
&\int d\vOmega \int dE\: \zeta(\rvec, \vOmega, E) \int d\vOmega' \int dE'\: \Sigma_s(\rvec, E \rightarrow E', \vOmega \cdot \vOmega') \zeta^{\dagger}(\rvec, \vOmega', E') 
\end{align*}
\end{itemize}
%
We now combine all three terms to get the adjoint form of the transport equation (!):
\begin{align*}
\langle\zeta^{\dagger}, H \zeta\rangle = \int_{V} dV \int_{4\pi} d\vOmega \int_0^{\infty} dE \: &\zeta (\rvec, \vOmega, E) \times \biggl[-\vOmega \cdot \nabla \zeta^{\dagger}(\vec{r}, E, \vOmega)  + \Sigma_t(\rvec,E) \zeta^{\dagger}(\vec{r}, E, \vOmega) \\&- \int d\vOmega' \int dE'\: \Sigma_s(\rvec, E \rightarrow E', \vOmega \cdot \vOmega') \zeta^{\dagger}(\rvec, \vOmega', E') \biggr]
\end{align*}
%
We can see that $\langle\zeta^{\dagger}, H \zeta\rangle = \langle\zeta, H^{\dagger} \zeta^{\dagger}\rangle$ will be satisfied if we define the adjoint operator as
\[
H^{\dagger} = -\vOmega \cdot \nabla  + \Sigma_t(\rvec,E) - \int d\vOmega' \int dE'\: \Sigma_s(\rvec, E \rightarrow E', \vOmega \cdot \vOmega')
\]
Where key differences are the negative in front of the streaming term and the reversal of energy direction in the scattering cross section.\\
We also need the boundary condition of zero \textit{outgoing} adjoint flux.

------------------------\\
Interpretation of adjoint flux as \textit{neutron importance}

Let's think about a problem with an external source: $H \psi = q_{ex}$, where $\psi$ meets the boundary conditions.\\
We'd like to find out the response of a detector with cross section $\Sigma_d$ and volume $V_d$ to that source: $R = V_d \int dE\: \Sigma_d(E) \int_{4\pi} \psi(\rvec_d, \vOmega, E)$. \\
\begin{itemize}
\item we'll set up the adjoint problem with the detector response as the source
\[H^{\dagger}\psi^{\dagger} = q_{ex}^{\dagger} \equiv V_d \Sigma_d \delta(\rvec - \rvec_d)\]
where $\psi^{\dagger}(\vec{r}_s, \vOmega, E) = 0$ for $\rvec$ on $\Gamma$ and $\vOmega \cdot \hat{n} > 0$.
%
\item We can multiply the initial equation (forward) by $\psi^{\dagger}$ and integrate over phase space. We can do the opposite/same to the adjoint equation. This gives
\begin{align*}
\langle\psi^{\dagger}, H\psi\rangle &= \langle\psi^{\dagger}, q_{ex}\rangle \quad \text{and}\\
\langle\psi, H^{\dagger} \psi^{\dagger}\rangle &= \underbrace{\langle\psi, V_d \Sigma_d \delta(\rvec - \rvec_d)\rangle }_R
\end{align*}
%
\item We can subtract the two and then apply the adjoint equivalence identity to see
\begin{align*}
\langle\psi^{\dagger}, H\psi\rangle - \langle\psi, H^{\dagger} \psi^{\dagger}\rangle &= \langle\psi^{\dagger}, q_{ex}\rangle - R \\
R &= \langle\psi^{\dagger}, q_{ex}\rangle
\end{align*}
\end{itemize}
%
So what does that mean? \\
It means that the detector response is the volume integral of the adjoint-weighted source distribution. \\
What does that mean? \\
The weighting of the adjoint flux directly corresponds to how influential a given source particle will be on the answer we're looking for (what we defined as the adjoint source).

This is a pretty powerful concept. \\
We can look at a delta source to clarify this a bit.
\begin{align*}
q_{ex} &= \delta(\rvec - \rvec_0) \delta(E - E_0) \delta(\vOmega \cdot \vOmega_0) \\
R &= \psi^{\dagger}(\rvec_0, \vOmega_0, E_0)
\end{align*}
It is clear here that the importance provides the detector response to particles produced at $(\rvec_0, \vOmega_0, E_0)$.

This also makes sense if we think through the differences in the equations:
\begin{itemize}
\item $-\vOmega \cdot \nabla$: by convention, $\vOmega \cdot \nabla \psi$ implies the \textit{out}flow of $\psi$ from an infinitesimal volume.\\
A neutron that exits $dV$ has a better chance to contribute to whatever you are measuring and so has a higher importance. \\
In the adjoint equation the streaming is therefore a \textit{gain} term (recall it's a loss in the forward, so the negative makes it production in the adjoint).
%
\item Scattering term: transposed $E'$ and $E$, $\vOmega'$ and $\vOmega$. \\
A neutron with energy $E$ and direction  $\vOmega$ has importance $\psi^{\dagger}(\rvec, \vOmega, E)$. \\
If it scatters into $E'$ and $\vOmega'$ it produces neutrons whose importance is $\psi^{\dagger}(\rvec, \vOmega', E')$.\\
Thus, the contribution to $\psi^{\dagger}(\rvec, \vOmega, E)$ from scattering is:
\[
\int_0^{\infty} dE' \int_{4\pi} d\vOmega' \:\Sigma_s(\rvec, E \rightarrow E', \vOmega \cdot \vOmega')\psi^{\dagger}(\rvec, \vOmega', E')
\]
\item Boundary condition: $\psi^{\dagger}(\rvec, \vOmega, E) = 0\:, \: \hat{n}\cdot \vOmega > 0$.\\
Neutrons exiting the problem configuration cannot scatter back and so cannot contribute to the detector inside. 
\end{itemize}

------------------------\\
The Adjoint Equation in \textbf{Multiplying Media}

We will call the fission operator $G$, where
\[
G \zeta = \frac{\chi(E)}{4\pi}\int_0^{\infty} dE' \: \nu(E') \Sigma_f(E') \int_{4\pi} d\vOmega' \psi(\rvec, \vOmega', E')
\]
To get the adjoint version, we can take the same approach as with the scattering operator to get
\[
G^{\dagger} \zeta^{\dagger} = \frac{\nu(E) \Sigma_f(E)}{4\pi}\int_0^{\infty} dE' \: \chi(E') \int_{4\pi} d\vOmega' \psi^{\dagger}(\rvec, \vOmega', E')
\]
We can combine this operator with $H$ to use the techniques in the preceding section to look at subcritical systems.\\
We replace $H$ and $H^{\dagger}$ with $H-G$ and $H^{\dagger}-G^{\dagger}$, respectively.\\

To look at the \textit{eigenvalue} form of the problem:
\[
H\psi - \frac{1}{k}G\psi = 0
\]
where $\psi$ is the fundamental solution mode, $\psi > 0$ for $\rvec \in V$, $\psi(\vec{r}_s, \vOmega, E) = 0$ for $\hat{n} \cdot \vOmega < 0$ with $\vec{r}_s \in \Gamma$.  
\[
H^{\dagger}\psi^{\dagger} - \frac{1}{k^{\dagger}}G^{\dagger}\psi^{\dagger} = 0
\]
Now $\psi^{\dagger}$ is the fundamental mode, $\psi^{\dagger} > 0$ for $\rvec \in V$, $\psi^{\dagger}(\vec{r}_s, \vOmega, E) = 0$ for $\hat{n} \cdot \vOmega > 0$ with $\vec{r}_s \in \Gamma$.  

Next, compare the two equations the way we did for response: multiply the forward equation by $\psi^{\dagger}$, integrate both over volume, and subtract
\begin{align*}
\langle\psi^{\dagger}, H\psi\rangle - \langle\psi, H^{\dagger} \psi^{\dagger}\rangle &- \frac{1}{k}\langle\psi^{\dagger}, G\psi\rangle + \frac{1}{k^{\dagger}}\langle\psi, G^{\dagger}\psi^{\dagger}\rangle = 0\\
\bigl(\frac{1}{k} - \frac{1}{k^{\dagger}} \bigr)\langle\psi^{\dagger}, G\psi\rangle  &= 0 \\
k &= k^{\dagger} \quad \text{because }G\psi > 0
\end{align*}

\clearpage----------------------------\\
\textbf{Reactivity} of a \textbf{Perturbed System}

Recall that reactivity is the measure of a system's deviation from critical: $\rho = \frac{k-1}{k}$.\\
We are interested in how small changes (perturbations) in a system will cause reactivity to change.\\
We can use the adjoint equation to help us.

We will denote small changes with a preceding $\delta$ and unperturbed values with a subscript $0$. Thus, for small changes in $k$ we can say 
\[
\delta \rho = \frac{\delta k}{k_0^2}
\]
For an unperturbed reactor we have these fundamental mode equations:
\begin{align*}
H_0\psi_0 &- \frac{1}{k_0}G_0\psi_0 = 0\\
H_0^{\dagger}\psi_0^{\dagger} &- \frac{1}{k_0^{\dagger}}G_0^{\dagger}\psi_0^{\dagger} = 0
\end{align*}
We can represent system perturbation by adding small quantities to the operators and fluxes (very small compared to things themselves). For the forward system we get
\[
(H_0 + \delta H)(\psi_0 + \delta \psi) - \frac{1}{k_0 + \delta k}(G_0 + \delta G)(\psi_0 + \delta \psi) = 0
\]
Because perturbations are small, we can ignore second order terms. Then
\begin{align*}
H_0 \psi_0 + \delta H \psi_0 &+ H_0\delta\psi - \bigl(\frac{1}{k_0 + \delta k}\bigr)\bigl(G_0\psi_0 + \delta G \psi_0 + G_0\delta\psi \bigr) \approx 0 \\
%
\frac{1}{k_0 + \delta k} &= \frac{1}{k_0(1 + \delta k/k_0)} \approx \frac{1}{k_0}\bigl(1 - \frac{\delta k}{k_0}\bigr) \\
\\
%
-\frac{\delta k}{k_0^2}G_0\psi_0 &\approx \underbrace{\bigl(H_0 - \frac{1}{k_0}G_0 \bigr)\psi_0}_{\text{fundamental mode}=0} + \bigl(\delta H - \frac{1}{k_0}\delta G\bigr)\psi_0  + \bigl(H_0 - \frac{1}{k_0}G_0 \bigr)\delta \psi
\end{align*}
To use inner product identities, we can multiply by $\psi_0^{\dagger}$ and integrate over phase space to get
\begin{equation}
-\frac{\delta k}{k_0^2} \langle \psi_0^{\dagger}, G_0\psi_0 \rangle \approx \langle \psi_0^{\dagger},\bigl(\delta H - \frac{1}{k_0}\delta G\bigr)\psi_0 \rangle + \langle \psi_0^{\dagger},\bigl(H_0 - \frac{1}{k_0}G_0 \bigr)\delta \psi \rangle
\label{eq:perturb}
\end{equation}

Next, we can use the adjoint identity (using a modified operator) and let $\zeta = \delta\psi$ and $\zeta^{\dagger} = \psi_0^{\dagger}$ we can see:
\begin{align*}
\langle\zeta^{\dagger}, H \zeta\rangle &= \langle\zeta, H^{\dagger} \zeta^{\dagger}\rangle \\
%
\langle \psi_0^{\dagger}, \bigl(H_0 - \frac{1}{k_0}G_0 \bigr)\delta \psi \rangle &= \langle \delta \psi,  \bigl(H_0^{\dagger} - \frac{1}{k_0^{\dagger}}G_0^{\dagger} \bigr)\psi_0^{\dagger} \rangle
\end{align*}
We know that the second part of the rhs is 0 because of the definition of the adjoint equation.\\
This means the lhs is also 0, we can can remove it from Eqn.~\ref{eq:perturb}.

With all of this, we can now say
\[
\frac{\delta k}{k_0^2} \approx \frac{\langle \psi_0^{\dagger},\bigl(\delta H - \frac{1}{k_0}\delta G\bigr)\psi_0 \rangle}{\langle \psi_0^{\dagger}, G_0\psi_0 \rangle}
\]
This means that if we know the two unperturbed solutions $\psi_0^{\dagger}$ and $\psi$ we can find the reactivity change for any perturbations.

\end{document}
